---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
download.file(url="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile="download.zip")
unzip("download.zip")
raw_data<-read.csv("activity.csv",stringsAsFactors = FALSE)
```


## What is mean total number of steps taken per day?
```{r}
per_day<-aggregate(x=raw_data$steps,
                   list(raw_data$date),
                   sum)
names(per_day)<-c("Date","Total_Steps")
mean_per_day<-mean(per_day$Total_Steps,na.rm=TRUE)
median_per_day<-mean(per_day$Total_Steps,na.rm=TRUE)
hist(per_day$Total_Steps, 
     xlab="Steps per day",
     ylab="Number of occurrences",
     main="Histogram of steps per day",
     col="white",
     border="black")
abline(v=mean_per_day,lty="dashed",col="red")
abline(v=median_per_day,lty="dashed",col="green")
text(x=mean_per_day,
     y=20,
     labels=paste("Mean value: ",
                  round(mean_per_day,digits=3),
                  "\n Median: ",
                  round(median_per_day,digits=3)),
     pos=4)
```

Mean is `r sprintf("%.3f",mean_per_day)` and median is `r sprintf("%.3f",median_per_day)`
```{r}
mean(per_day$Total_Steps,na.rm=TRUE)
```
Median:
```{r}
median(per_day$Total_Steps,na.rm=TRUE)
```


## What is the average daily activity pattern?
```{r}
per_5_min<-aggregate(x=raw_data$steps,list(raw_data$interval),mean,na.rm=TRUE)
names(per_5_min)<-c("min_interval","Total_Steps")
plot(x=per_5_min$min_interval,
     y=per_5_min$Total_Steps,
     type="l",
     xlab="5 min interval",
     ylab="Mean number of steps")
max_interval=per_5_min$min_interval[
  which.max(per_5_min$Total_Steps)]
max_value=max(per_5_min$Total_Steps)
abline(v=max_interval,
       col="blue",
       lty="dashed")
text(x=max_interval,
     y=max_value-50,
     labels=paste("Max value: ",
                  round(max_value,digits=3),
                  "\n At interval: ",
                  max_interval),
     pos=4)
```

Max average number of steps, `r sprintf("%.3f",max_value)`, happens at 5 min interval `r max_interval`

## Imputing missing values

Number of missing values in the set:
```{r}
sum(is.na(raw_data$steps))
```

Replace missing values with mean for that interval
```{r}
na_vector=is.na(raw_data$steps)
replace_missing<-raw_data
replace_missing$interval=as.factor(replace_missing$interval)
per_5_min$min_interval=as.factor(per_5_min$min_interval)
#Replace NA with mean from same interval
replace_missing$steps[na_vector]=mean(
  per_5_min$Total_Steps
  [replace_missing$interval[na_vector]])
sum(is.na(replace_missing$steps))
```

Now create the histogram with the replaced dataset

```{r}
per_day<-aggregate(x=replace_missing$steps,
                   list(replace_missing$date),
                   sum)
names(per_day)<-c("Date","Total_Steps")
#We can ommit the na.rm=TRUE, given that NA are replaced
mean_per_day<-mean(per_day$Total_Steps)
median_per_day<-mean(per_day$Total_Steps)
hist(per_day$Total_Steps, 
     xlab="Steps per day",
     ylab="Number of occurrences",
     main="Histogram of steps per day",
     col="white",
     border="black")
abline(v=mean_per_day,lty="dashed",col="red")
abline(v=median_per_day,lty="dashed",col="green")
text(x=mean_per_day,
     y=20,
     labels=paste("Mean value: ",
                  round(mean_per_day,digits=3),
                  "\n Median: ",
                  round(median_per_day,digits=3)),
     pos=4)
```

Mean is `r sprintf("%.3f",mean_per_day)` and median is `r sprintf("%.3f",median_per_day)`

Mean:
```{r}
mean(per_day$Total_Steps,na.rm=TRUE)
```
Median:
```{r}
median(per_day$Total_Steps,na.rm=TRUE)
```
As we are replacing the missing values with the average value per minute interval, there is no difference between the two sets (the one with missing values and the one with replaced values)

## Are there differences in activity patterns between weekdays and weekends?

```{r}
library(lubridate)
replace_missing$Date<-ymd(replace_missing$date)
#Use wday to detect if is Weekend (1 or 6) or Weekday
replace_missing$day_type<-as.numeric(wday(replace_missing$date)==1 | wday(replace_missing$date)==7)+1
replace_missing$day_type=as.factor(replace_missing$day_type)
levels(replace_missing$day_type)<-c("weekday","weekend")
per_5_min_type<-aggregate(x=replace_missing$steps,
                          list(replace_missing$interval,
                               replace_missing$day_type),
                          mean)
names(per_5_min_type)<-c("min_interval","day_type","Total_steps")
#De-factor min_interval so that lattice plot considers it as numer
per_5_min_type$min_interval<-as.numeric(levels(per_5_min_type$min_interval))
library(lattice)
xyplot(Total_steps ~ min_interval|day_type,
       data=per_5_min_type,
       layout=c(1,2),
       type="l",
       xlab="minute interval",
       ylab="mean number of steps",
       main="Mean number of steps per interval and type of day"
       )
```

We do observe that average steps on weekdays peak on the morning and here and there on the afternoon. However on weekends, there are numerous peaks of activity throughout the day
